Strong lensing selection effects

### 2022-07-10 paper/seleff.tex
glafic tests. Section 2

The structure of this work is the following.
\Sref{sect:basics} introduces the basics of gravitational lensing.
In \Sref{sect:indlenses} we study individual lens systems, to look for trends in the lensing cross-section with various lens and source properties.
In \Sref{sect:lenspop} we describe our simulations of lens populations.
In \Sref{sect:results} we show the results of our analysis on the simulated lens populations.
We discuss the results in \Sref{sect:discuss} and draw conclusions in \Sref{sect:concl}.

@@

The lensing properties of an object with respect to a source depend solely on its dimensionless surface mass density distribution $\kappa(\boldsymbol\theta)$ (also referred to as the convergence), 
% ratio between its surface mass density distribution and the critical surface mass density for lensing:
\begin{equation}
\kappa(\boldsymbol\theta) = \dfrac{\Sigma(\boldsymbol\theta)}{\Sigma_{\mathrm{cs}}},
\end{equation}
which is the ratio between the projected surface mass density of the lens and the critical surface mass density for lensing.
The latter quantity is defined as
%The quantity $\Sigma_{\matrhrm{cs}}$ is the critical surface mass density for lensing, defined as
\begin{equation}
\Sigma_{cr} = \dfrac{c^2D_s}{4\pi G D_d D_{ds}},
\end{equation}
where $c$ is the speed of light, $G$ the gravitational constant, and $D_d$, $D_s$, and $D_{ds}$ are the angular diameter distances between the observer and the lens, the observer and the source, and the lens and the source, respectively.

@@

A very important quantity for describing the strength of a strong lens is the Einstein radius, $\tein$. For an axisymmetric lens, this is the radius corresponding to the solution of \Eref{eq:lenseq} for a source placed at the same angular position as the lens centre ($\beta=0$).
The circle with radius equal to $\tein$ is known as the tangential critical curve. Images that form here have infinite magnification in the tangential direction.

It can be shown that $\tein$ satisfies the following condition,
\begin{equation}\label{eq:criteq}
\bar{\kappa}(<\tein) = 1,
\end{equation}
where $\bar{\kappa}(<\theta)$ is the average surface mass density within a radius equal to $\theta$:
\begin{equation}
\bar{\kappa}(<\theta) \equiv \frac{2}{\theta^2}\int_0^{\theta}d\theta' \kappa(\theta')\theta'.
\end{equation}
Based on \Eref{eq:criteq},
%The average surface mass density within $\tein$ is equal to one.
%For this reason, 
we generalise the definition of $\tein$ to the non-axisymmetric case as the circularised radius of the isodensity contour that encloses an average surface mass density equal to $\Sigma_{\mathrm{cr}}$.

Axisymmetric lenses with a density profile of the kind considered in section \ref{ssec:profile} typically produce either one or three images of a point source.
\Fref{fig:scheme} helps to visualise this property.
%\Fref{fig:scheme}, which shows the quantity $\theta - \alpha(\theta)$ as a function of the position in the image plane $\theta$, helps to visualise this property. 
Plotted in \Fref{fig:scheme} is the quantity $\theta - \alpha(\theta)$ as a function of the position in the image plane $\theta$, for the same lens models considered in \Fref{fig:kappa}.
According to \Eref{eq:1dlenseq}, images form at the locations where this quantity equals the position of the source. 
Therefore, given a source position $\beta$, the number of images and their location can be determined by drawing a horizontal line at the value $\beta$ on the vertical axis, and finding the points where this line intersects the $\theta - \alpha(\theta)$ curves.

@@

\begin{figure}
\includegraphics[width=\columnwidth]{composite_fixedap_scheme.eps}
\caption{
The lens equation of axisymmetric lenses. Solid lines: right-hand side of \Eref{eq:1dlenseq} for the five lenses with density profile plotted in \Fref{fig:kappa}. 
Given a source at position $\beta$, its lensed images form at the values of $\theta$ where the solid line intersects a horizontal line located at $\beta$ on the vertical axis. 
Dotted lines: positions of the radial caustics. Sources located within the radial caustic of a given lens produce three images.
\label{fig:scheme}
}
\end{figure}
%The slope of the $\theta - \alpha(\theta)$ curve is the inverse of the magnification of an image in the radial direction. 
%This can be understood by noting that, according to \Eref{eq:1dlenseq}, the derivative of $\theta - \alpha(\theta)$ is equal to $d\beta/d\theta$, which quantifies the ratio between an infinitesimal displacement in the radial direction in the source plane and the corresponding displacement in the image plane.
%Stationary points of the $\theta - \alpha(\theta)$ curve, therefore, are points of infinite magnification in the radial direction. For a given axisymmetric lens, these points describe a circle known as the radial critical curve. 
%The corresponding curve in the source plane, the radius of which is given by the value of $\theta - \alpha(\theta)$ at the stationary point, is known as the radial caustic.
%As can be seen in \Fref{fig:scheme}, sources that lie within the radial caustic are strongly lensed into three images, while sources that lie outside of it produce only one image.

>>>>

The structure of this work is the following.
In \Sref{sect:basics} we introduce the basics of gravitational lensing.
In \Sref{sect:indlenses} we study individual lens systems, to look for trends in the lensing cross-section with various lens and source properties.
In \Sref{sect:lenspop} we describe our simulations of lens populations.
In \Sref{sect:results} we show the results of our analysis on the simulated lens populations.
We discuss the results in \Sref{sect:discuss} and draw conclusions in \Sref{sect:concl}.

@@

The lensing properties of an object with respect to a source depend solely on its dimensionless surface mass density distribution $\kappa(\boldsymbol\theta)$ (also referred to as the convergence). 
This is the ratio between the surface mass density and the critical surface mass density for lensing:
\begin{equation}
\kappa(\boldsymbol\theta) = \dfrac{\Sigma(\boldsymbol\theta)}{\Sigma_{\mathrm{cs}}}.
\end{equation}
%which is the ratio between the projected surface mass density of the lens and the critical surface mass density for lensing.
The latter quantity is defined as
%The quantity $\Sigma_{\matrhrm{cs}}$ is the critical surface mass density for lensing, defined as
\begin{equation}
\Sigma_{cr} = \dfrac{c^2D_s}{4\pi G D_d D_{ds}},
\end{equation}
where $c$ is the speed of light, $G$ the gravitational constant, and $D_d$, $D_s$, and $D_{ds}$ are the angular diameter distances between the observer and the lens, the observer and the source, and the lens and the source, respectively.

@@

A very important quantity for describing the strength of a strong lens is the Einstein radius, $\tein$. For an axisymmetric lens, this is the radius corresponding to the solution of \Eref{eq:lenseq} for a source placed at the same angular position as the lens centre ($\beta=0$).
The circle with radius equal to $\tein$ is known as the tangential critical curve. Images that form here have infinite magnification in the tangential direction.
%
It can be shown that $\tein$ satisfies the following condition,
\begin{equation}\label{eq:criteq}
\bar{\kappa}(<\tein) = 1,
\end{equation}
where $\bar{\kappa}(<\theta)$ is the average surface mass density within a radius equal to $\theta$:
\begin{equation}
\bar{\kappa}(<\theta) \equiv \frac{2}{\theta^2}\int_0^{\theta}d\theta' \kappa(\theta')\theta'.
\end{equation}
%Based on \Eref{eq:criteq},
%we generalise the definition of $\tein$ to the non-axisymmetric case as the circularised radius of the isodensity contour that encloses an average surface mass density equal to $\Sigma_{\mathrm{cr}}$.

Axisymmetric lenses of the kind considered in this work 
%Axisymmetric lenses with a density profile of the kind considered in section \ref{ssec:profile} 
typically produce either one or three images of a point source.
\Fref{fig:scheme} helps to visualise this property.
%\Fref{fig:scheme}, which shows the quantity $\theta - \alpha(\theta)$ as a function of the position in the image plane $\theta$, helps to visualise this property. 
Plotted in \Fref{fig:scheme} is the quantity $\theta - \alpha(\theta)$ as a function of the position in the image plane $\theta$, for a few lens models.
%for the same lens models considered in \Fref{fig:kappa}.
According to \Eref{eq:1dlenseq}, images form at the locations where this quantity equals the position of the source. 
Therefore, given a source position $\beta$, the number of images and their location can be determined by drawing a horizontal line at the value $\beta$ on the vertical axis, and finding the points where this line intersects the $\theta - \alpha(\theta)$ curve.

@@

\begin{figure}
\includegraphics[width=\columnwidth]{composite_fixedap_scheme.eps}
\caption{
The lens equation of axisymmetric lenses. Solid lines: right-hand side of \Eref{eq:1dlenseq} for five lenses with different density profiles. 
Given a source at position $\beta$, its lensed images form at the values of $\theta$ where the solid line intersects a horizontal line located at $\beta$ on the vertical axis. 
Dotted lines: positions of the radial caustics. Sources located within the radial caustic of a given lens produce three images.
The lens models used in this simulations consist of a stellar component and a dark matter halo, as described in section \ref{ssec:profile}. Their density profile is plotted in \Fref{fig:kappa}.
\label{fig:scheme}
}
\end{figure}
%The slope of the $\theta - \alpha(\theta)$ curve is the inverse of the magnification of an image in the radial direction. 
%This can be understood by noting that, according to \Eref{eq:1dlenseq}, the derivative of $\theta - \alpha(\theta)$ is equal to $d\beta/d\theta$, which quantifies the ratio between an infinitesimal displacement in the radial direction in the source plane and the corresponding displacement in the image plane.
%Stationary points of the $\theta - \alpha(\theta)$ curve, therefore, are points of infinite magnification in the radial direction. For a given axisymmetric lens, these points describe a circle known as the radial critical curve. 
%The corresponding curve in the source plane, the radius of which is given by the value of $\theta - \alpha(\theta)$ at the stationary point, is known as the radial caustic.
%As can be seen in \Fref{fig:scheme}, sources that lie within the radial caustic are strongly lensed into three images, while sources that lie outside of it produce only one image.
